\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-24 paper.tex

\section{ARGs and local trees}
\label{sec-ARG-and-local-trees}
% (We use ``local'' rather than
% ``marginal''
% % could list loads here, but not much point.
% trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
% to avoid the potential confusion with the
% statistical concept of marginal distributions.)
A fundamental property of genetics is that a
% TODO is this (genome) clarification helpful or confusing?
given DNA nucleotide is inherited from exactly one parent (genome),
both at an organismal and cell-by-cell level
(Appendix~\ref{sec-cell-lineages-and-args}).
These chains of single-parent inheritance over time give rise,
by definition, to a tree structure.
As a result of recombination, adjacent nucleotides can have
different chains of inheritance, and
an ARG encodes the entire ensemble of local trees along the
genome for a given set of sample genomes.
The relationship between an ARG and its corresponding
local trees is subtle and important. To make the following
discussions precise, we will define a concrete combinatorial encoding
for local trees and an explicit procedure for constructing them.

Oriented trees provide a
convenient formalism to capture these relationships in a well-defined
combinatorial object.
% There's no need for a comma
Let $\pi_1\dots\pi_n$ be a sequence of integers, such that $\pi_u$
denotes the parent of node $u$, and $\pi_u = 0$ if $u$ is a
root~\cite[p.\ 461]{knuth11combinatorial}.
This encoding is particularly useful to describe
evolutionary trees because parent-child relationships are
important but the ordering of children at a node is
not~\citep{kelleher2013coalescent,kelleher2014coalescent,
kelleher2016efficient}.
Thus, for a given gARG with nodes $\{1, \dots, n\}$ and
edges $E$, we recover the local tree as follows.
We begin by
setting $\pi_u = 0$ for each $1 \leq u \leq n$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq 0$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).
Note that the trees for an ARG are ``sparse''~\citep{kelleher2016efficient},
because many ancestral nodes will not be reachable from the
samples at a given position.

This combinatorial approach provides at least
one novel insight into the relationship between ARGs and local trees.
If we are given the sequence of local trees for a gARG
encoded as an oriented tree along with the genome interval
covered by each tree, then we can recover precisely the same ARG.
% I guess we should be a bit more formal here?
Thus, under this definition, there
is a one-to-one correspondence between an ARG and
the sequence of local trees that it encodes.

This is not the prevailing view, however.
\cite{kuhner2017consensus} argue that the
``interval-tree'' representation of
an ARG (the local trees and the genome intervals they cover)
``does not contain all of the information in the underlying ARG: it lacks the
number of recombinations occuring at each site, the times at which
recombinations occurred, and the specific sequences involved as recombination
partners.''
\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
% Could cite some more? E.g. KwARG paper says
% "We note that ARGs contain more information than local trees"
% If even Ana and Jotun are wrong about this it then it must be
% significant! ;)
These statements that an ARG
contains more information than its local trees is only true
if we use a representation of the local trees that discards
information about those local trees that can be derived from the ARG.

There are two properties of how evolutionary trees are conventionally
represented that lead to this
% Yes, OK, we can rephrase. It is a misunderstanding though
misunderstanding of the relationship between local trees and an ARG.
Firstly, the internal nodes of evolutionary trees are usually
considered to be \emph{unlabelled}.
Defining ARG nodes as integers and using the oriented tree encoding
explicitly labels internal nodes, and makes the relationship between
tree and ARG nodes explicit.

[WIP]

because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes may likewise correspond to anonymous hypothetical
ancestors, but if we do not label the tree nodes we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Besides having labelled nodes, the local trees must contain
unary nodes marking recombinations
and other processes that lead to lineages passing through a given
ARG node without coalescence in the local trees
(see Section~\ref{sec-locally-unary-edges}).
If the local trees contain labelled internal nodes, they precisely
describe the corresponding path through the graph for each
genome interval and there is exactly as
much information in them as in an ARG.

These properties have interesting consequences when we
consider the inference strategy of estimating local trees
% TODO rephrase this. Ana doesn't like the way it implies that the
% trees are independent, because info from mutations outside the
% window are shared. However, each tree *inference* is independent,
% whether information is used multiple times in different trees or not
independently within genome segments,
and subsequently stitching them together into an ARG.
Two recent methods take this approach,
\relate~\citep{speidel2019method} and
\espalier~\citep{rasmussen2022espalier},
and have different strategies for reconciling the local trees into an ARG.
\relate\ first infers local tree topologies, and then identifies
edges that persist from tree to tree by reasoning about the sets of samples
subtended. It does not attempt to be exhaustive.
\espalier, on the other hand, is specifically interested
in the details of recombination events (among a small number of samples)
and therefore attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.

\section{Locally unary nodes}
\label{sec-locally-unary-edges}
% TODO needs intro, giving some bit of warm-up for the reader.
We can define the arity of a node as the number of child nodes it has.
There is an
important distinction between the number of children a node has
in the graph and the number of children it has in the local trees,
because a (graph) child node is only a child in the local tree for position
% FIXME state this formally, as both Yan and Jere misunderstood here.
$x$ if $x$ is in the inheritance interval for the corresponding edge.
Thus, for a given node $u$ and a genome position $x$ the arity
of $u$ in that local tree is less than or equal to the graph arity of
$u$. The arity
of a node in the local trees can therefore change as we move along
the genome.

This distinction between graph and local tree arity is mainly
of interest when we consider \emph{unary} nodes, those that have
a single child.
Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
\noderef{g} and \noderef{h} are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
On the other hand, node \noderef{f} has two children
in the graph, but is binary only
in the local tree covering the interval $[2, 7)$,
representing the coalescence of samples \noderef{a} and \noderef{c}
in this genome region. Over the interval $[0, 2)$ no coalescence occurs,
but we still record the fact that genome \noderef{c} inherits from \noderef{f}
in the local tree. Thus, \noderef{c} has a single child in this
interval and it is unary in this local tree.
% TODO note this is CA but not coalescent, and refer to where we discuss this.
In another example, \noderef{e} is binary in the graph, but is
unary in each of the local trees in which it is present.
This is because \noderef{a} and \noderef{c} are both recombinant
offspring of \noderef{e} and \noderef{f}, and,
in this example,
\noderef{a} inherits from
\noderef{e} over the interval $[0, 2)$,
and \noderef{c} inherits from \noderef{e} over $[7, 10)$.
The intervals over which
we know that a node inherits from another, but where there
is no local coalescence, are marked by unary nodes in the
local trees.

Outside of the context of simulation, ARG-unary nodes are
only likely to occur in longitudinal datasets where genetic
data is sampled at many timepoints and recombination
is rare (e.g.~SARS-CoV-2; see Discussion).
\emph{Locally} unary nodes, those which have one child
in one or more local trees, are a common and important feature
of many different types of ARG, and a recurring topic
in this article. Without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them.

% FIXME: move to top of section
Nodes with one child are not a standard feature of evolutionary trees
or the ARG literature
(although \citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.'').
We are usually interested in
nodes that have at least two children because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions; see Appendix \ref{sec-cell-lineages-and-args})
and there is usually
little information about these intermediates, and hence limited utility
in modelling them.
As we see in the next section, however, there is a range of
different ways in which ARG nodes can be locally unary. Furthermore,
several current ARG inference methods infer locally unary nodes
(see Section~\ref{sec-precision}).

>>>>

\section{ARGs and local trees}
\label{sec-ARG-and-local-trees}
% (We use ``local'' rather than
% ``marginal''
% % could list loads here, but not much point.
% trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
% to avoid the potential confusion with the
% statistical concept of marginal distributions.)
The relationship between an ARG and its corresponding
local trees is subtle and important.
A fundamental property of genetics is that a
given DNA nucleotide is inherited from exactly one parent genome,
both at an organismal and cell-by-cell level
(Appendix~\ref{sec-cell-lineages-and-args}).
These chains of single-parent inheritance over time give rise,
by definition, to a tree structure.
As a result of recombination, adjacent nucleotides can have
different chains of inheritance, and
an ARG encodes the entire ensemble of local trees along the
% Said "sample of genomes" originally, but using term "genome" in two
% different ways in the same sentence then.
genome for a given sample.
% This is for Yan - to me this is a given, but maybe some people feel
% the following is overly formal?
Precisely defining the process by which local trees are extracted
from an ARG is essential to our understanding of how they
are related, and we require a concrete mathematical structure
to describe the local trees.

Oriented trees provide a
convenient formalism to capture these parent-child relationships
in a well-defined combinatorial object.
% There's no need for a comma
Let $\pi_1\dots\pi_n$ be a sequence of integers, such that $\pi_u$
denotes the parent of node $u$, and $\pi_u = 0$ if $u$ is a
root~\cite[p.\ 461]{knuth11combinatorial}.
This encoding is particularly useful to describe
evolutionary trees because parent-child relationships are
important but the ordering of children at a node is
not~\citep{kelleher2013coalescent,kelleher2014coalescent,
kelleher2016efficient}.
Thus, for a given gARG with nodes $\{1, \dots, n\}$ and
edges $E$ (Section~\ref{sec-gARG}), we recover the local tree
at position $x$ as follows.
We begin by setting $\pi_u = 0$ for each $1 \leq u \leq n$.
Then, for each sample node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq 0$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).
Note that the local trees for an ARG are ``sparse''~\citep{kelleher2016efficient},
because many ancestral nodes will not be reachable from the
samples at a given position.

This combinatorial approach provides at least
one novel insight into the relationship between ARGs and local trees.
If we are given the sequence of local trees for a gARG
encoded as an oriented tree along with the genome interval
covered by each tree, then we can recover precisely the same ARG.
% We do this site-by-site for simplicity, but hopefully it's obvious
% that this isn't necessary?
More formally, suppose we are given the local tree $\pi^x_1\dots\pi^x_n$
for each nucleotide position $1 \leq x \leq L$ on a genome of length $L$.
Then, the edges of the ``local ARG'' for this tree is given by
$E^x = \{(u, \pi_u^x, \{x\} \mid \pi_u^x \neq 0\}$. Since the ARG
edges are defined by $(c, p, I)$ tuples, where the set $I$ defines
the positions over which node $c$ inherits from parent $p$, we can
then simply combine the ``local ARGs'' for each position $x$
to recover precisely the same set of edges as the original ARG.
Thus, under this definition, there
is a one-to-one correspondence between an ARG and
the sequence of local trees that it encodes.

This is not the prevailing view, however.
\cite{kuhner2017consensus} argue that the
``interval-tree'' representation of
an ARG (the local trees and the genome intervals they cover)
``does not contain all of the information in the underlying ARG: it lacks the
number of recombinations occuring at each site, the times at which
recombinations occurred, and the specific sequences involved as recombination
partners.''
\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
% Could cite some more? E.g. KwARG paper says
% "We note that ARGs contain more information than local trees"
% If even Ana and Jotun are wrong about this it then it must be
% significant! ;)
These statements that an ARG
contains more information than its local trees are only true
if we use a representation of the local trees that discards
information about those local trees that can be derived from the ARG.

There are two properties of how evolutionary trees are conventionally
represented that lead to this
% We can rephrase... suggestions?
disagreement about the relationship between local trees and an ARG.
Firstly, the internal nodes of evolutionary trees are usually
considered to be \emph{unlabelled}.
Evolutionary trees usually regard internal nodes as unlabelled
because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes may likewise correspond to anonymous hypothetical
ancestors, but if we do not label the tree nodes we lose the fact that
the \emph{same} ancestors sometimes persist across multiple trees.
Defining ARG nodes as integers and using the oriented tree encoding
explicitly labels internal nodes, and makes the relationship between
tree and ARG nodes clear.

The second property of how evolutionary trees
are conventionally represented that is unhelpful in the context of ARGs is their
focus on branching points (coalescences), i.e.\
nodes that have two or more children.
As the introductory paragraph of this section emphasised,
parent-child relationships are what fundamentally define a tree,
and branching points can be seen as incidental. This is reflected
by the oriented tree encoding which simply stores the local
parent-child relationships, and does not, for example,
directly tell us how many children a particular node has.
The local tree at a given position records the \emph{path} through
the ARG, and if we discard sections of that path that are not
branching points in the local tree, we lose information about the ARG.
We return to this important point in the following two sections.

It is important that we make the distinction here between the local
trees that we can derive from a known ARG (as just discussed),
and the ARGs that we can derive from a sequence of \emph{estimated}
local trees.
The recent
\espalier~\citep{rasmussen2022espalier} ARG inference method
is illustrative in this context.
\espalier\ works by splitting an input sequence alignment into
segments that are assumed to be non-recombining. Within
each segment, an initial local tree is estimated using
standard phylogenetic methods. By necessity, these local
trees will contain internal nodes that are unlabelled and
consist only of branching points: there is no information
shared between the independent tree estimation steps
across segments. Part of the task of stitching
these trees together
into an ARG is then, essentially, to generate labels for
the internal nodes, and decide which nodes persist
across multiple local trees.
\espalier\ is specifically interested
in the details of recombination events,
and also attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination between these partially reconciled local trees.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.
% I don't really know if this is true, it's just a way of wrapping the
% section up by saying that other things we thought were settled might
% be worth thinking about again.
The combinatorial formulation of ARGs and local trees provided here
may help clarify these fundamental questions.

\section{Locally unary nodes}
\label{sec-locally-unary-edges}
As discussed in the previous section, the local tree at a given position
$x$ is best seen as the path through the ARG at that position, defined
by the oriented tree $\pi^x_1\dots\pi^x_n$. This path does not directly
contain information about branching points, and so it is helpful
to define the ``arity''.
The ``local arity'' of a node is the number of children it has
in the local tree at position $x$, i.e., $a^x_u = |\{v : \pi^x_v = u\}|$
for each $1 \leq u \leq n$. The ``ARG arity'' of a node $u$ is the
number of children it has in the graph topology, i.e.,
$a_u = |\{v: (v, u, I) \in E\}|$. Thus, the local arity is less than
or equal to the ARG arity (more precisely, $0 \leq a^x_u \leq a_u$),
and the local arity of a node may change as we move along the genome.

This distinction between ARG and local tree arity is mainly
of interest when we consider \emph{unary} nodes, those that have
a single child.
Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
\noderef{g} and \noderef{h}
are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
On the other hand, node \noderef{f} has two children
in the graph, but is binary only
in the local tree covering the interval $[2, 7)$,
representing the coalescence of samples \noderef{a} and \noderef{c}
in this genome region. Over the interval $[0, 2)$ no coalescence occurs,
but we still record the fact that genome \noderef{c} inherits from \noderef{f}
in the local tree. Thus, \noderef{c} has a single child in this
interval and it is unary in this local tree.
% TODO note this is CA but not coalescent, and refer to where we discuss this.
In another example, \noderef{e} is binary in the graph, but is
unary in each of the local trees in which it is present.
This is because \noderef{a} and \noderef{c} are both recombinant
offspring of \noderef{e} and \noderef{f}, and,
in this example,
\noderef{a} inherits from
\noderef{e} over the interval $[0, 2)$,
and \noderef{c} inherits from \noderef{e} over $[7, 10)$.
The intervals over which
we know that a node inherits from another, but where there
is no local coalescence, are marked by unary nodes in the
local trees.

Outside of the context of simulation, ARG-unary nodes are
only likely to occur in longitudinal datasets where genetic
data is sampled at many timepoints and recombination
is rare (e.g.~SARS-CoV-2; see Discussion).
\emph{Locally} unary nodes, those which have one child
in one or more local trees, are a common and important feature
of many different types of ARG.
As discussed in the previous section,
without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them.

Nodes with one child are not a standard feature of evolutionary trees
or the ARG literature
(although \citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.'').
We are usually interested in
nodes that have at least two children because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions; see Appendix \ref{sec-cell-lineages-and-args})
and there is usually
little information about these intermediates, and hence limited utility
in modelling them.
As we see in the next section, however, there is a range of
different ways in which ARG nodes can be locally unary. Furthermore,
several inference methods infer ARGs containing locally unary nodes
(see Section~\ref{sec-precision}).
